Dish: Distribution-oblivious Shell Scripting

### 2019-11-24 pldi/draft.tex
Update background and overview

This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii) a high-level overview of \sys's design and implementation.

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
%\begin{wrapfigure}{r}{0.06\textwidth}
%  \vspace{-12pt}
%  % \begin{center}
%    % \includegraphics[width=0.48\textwidth]{gull}
%    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
%  % \end{center}
%  \vspace{-30pt}
%\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  ls -l | wc -l                          [($$p_2$$)]
\end{lstlisting}
\smallskip
Commands can be tuned by two mechanisms, namely com\-mand-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  after it expands special characters such as \ttt{~} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

% TODO: Talk about security primitives?
Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
  files are simply persistent streams that can be manipulated by commands or pipelines.
% \begin{wrapfigure}{r}{0.15\textwidth}
%   \vspace{-15pt}
%   % \begin{center}
%     % \includegraphics[width=0.48\textwidth]{gull}
%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
%   % \end{center}
%   \vspace{-25pt}
% \end{wrapfigure}
Reading a file or a set of files generates a stream; 
  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
Files are named, in the sense that they correspond to a global identifier---for example, \ttt{/x/y} corresponds to a persistent stream.
\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).

% At a high level, POSIX defines several file types:
% (-) regular files, storing lines of text or binary data,
% (d) directories, grouping together multiple other files,
% (l) symbolic links, pointing to other files,
% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  ls | sed 's/1/2/' | > out.t 2> err.t    [($$p_3$$)]
\end{lstlisting}
\smallskip

\noindent
Special files are particularly interesting because they are resolved to OS structures.
One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
% Here are a few examples:
% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.

A small but powerful set of features is concerned with stream manipulation.
Some of these features are built into the shell;
  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
% \begin{wrapfigure}{r}{0.15\textwidth}
%   \vspace{-15pt}
%   % \begin{center}
%     % \includegraphics[width=0.48\textwidth]{gull}
%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
%   % \end{center}
%   \vspace{-25pt}
% \end{wrapfigure}
As described before, the file can be a FIFO dubbing as input to another command.
More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
% TODO: talk about \ttt{xargs}?

\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  cat * | tee >(wc -l > l.t) | zip > a.z   [($$p_4$$)]
\end{lstlisting}
\smallskip

@@

\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  read | { calc | square; } >/dev/fd/0     [($$p_5$$)]
\end{lstlisting}
\smallskip

@@

\unix pipelines are surprisingly powerful, a feature that \sys exploits by re-targeting shell scripts during its rewriting passes that target shell.

@@

\subsection{\sys Design Overview}

As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
Intuitively,
 (i) the parallel-composition operator exposes task parallelism
% FIXME \kk{I have named that operator level parallelism, which one do we prefer?},
 (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
 (iii) the pipe operator is candidate for both task and data parallelism, and
% \kk{How does it enable data parallelism?} \nv{added "candidate", helps?}
 (iv)  other composition operators can be viewed as barrier candidates.
To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.

% However, understanding the shell's primitives is not enough
\sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
This challenge can be broken down into two parts.
First, there is a need to understand the commands already in the shell---that is, built-ins that come with any shell and support its core functions.
One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes.
This set is already useful enough to allow the composition of practical pipelines~\sx{eval}.

The second part is expressing such characteristics extensions that fall outside the built-in set, allowing new commands to leverage \sys's power.
Commands are arbitrary programs, but still conform to the shell's pipeline interface.
To solve this part, \sys leverages the prior study to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
Using this DSL, developers of commands (not users of the shell) can quickly and easily capture the class of the command.
A command developer can only annotate the default behavior;
  if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.

Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
\sys addresses this challenge with a series of rewriting passes that perform dataflow analysis for identifying  distributable regions.
It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.
These transformations are complicated by the shell's pervasive use of dynamic features---\eg environment variables, file input sizes, \etc

Finally, \sys needs to schedule and execute the distributed dataflow graph.
To achieve this, \sys's runtime component needs to solve several challenges~\sx{impl}.
First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.

Second, \sys needs to fill-in planing details left blank by the rewriting pass.
This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
Planner activators achieves the latest possible binding, when critical runtime information is available.
% FIXME \kk{I am not exactly sure I understand that :) } \nv{which part?}

The next few sections discuss \sys's design and implementation details.
They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.

>>>>

This section presents (i) important background on shell pipeline composition~\sx{bg:pipelines}, and (ii) a high-level overview of \sys's design and implementation~\sx{bg:overview}.

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed \emph{commands}, can also be developed in different programming languages, as long as they conform to a common interface.
%\begin{wrapfigure}{r}{0.06\textwidth}
%  \vspace{-12pt}
%  % \begin{center}
%    % \includegraphics[width=0.48\textwidth]{gull}
%    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
%  % \end{center}
%  \vspace{-30pt}
%\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  ls -l | wc -l                           ($$(p_2)$$)
\end{lstlisting}
\smallskip
Commands can be tuned by two mechanisms: com\-mand-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  after it expands special characters such as \ttt{\textasciitilde{}} and \ttt{*}, it leaves parsing and evaluation entirely up to every command.
Environment variables carry more general information about the surrounding system---\eg search paths, system defaults, parsing switches \etc 
They are organized as a map from names to values, both of which are strings, and can be evaluated at any point in the program---\eg as commands, flags, or inputs. % a shell feature termed parameter \emph{expansion}.

% TODO: Talk about security primitives?
Streams are a key abstraction in \unix, and go hand-in-hand with another \unix abstraction: \emph{files}.
Files are persistent streams that can be manipulated by individual commands or pipelines:
% \begin{wrapfigure}{r}{0.15\textwidth}
%   \vspace{-15pt}
%   % \begin{center}
%     % \includegraphics[width=0.48\textwidth]{gull}
%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
%   % \end{center}
%   \vspace{-25pt}
% \end{wrapfigure}
  reading a (set of) file(s) generates a stream; 
  conversely, redirecting to a file identifier generates a file.
% streams can be trivially redirected to the file system---by appending with the file identifier.
Files are named, corresponding to global identifiers---\eg \ttt{/x/y} corresponds to a persistent stream at the location \ttt{/y} in \ttt{/x}.
\unix exposes a few different types of files---\eg directories, symbolic links, and named pipes (or FIFOs). % , and special-character devices.
As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).

% At a high level, POSIX defines several file types:
% (-) regular files, storing lines of text or binary data,
% (d) directories, grouping together multiple other files,
% (l) symbolic links, pointing to other files,
% (p) named pipes or FIFOs, communication primitives that can be named and manipulated like files,
% (s) domain sockets or DSes, full-duplex communication primitives that support passing file descriptors, and
% (b or c) special files, interfaces to device drivers or low-level abstractions, presented as ordinary files.
\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  ls | sed 's/1/2/' | > out.t 2> err.t    ($$(p_3)$$)
\end{lstlisting}
\smallskip

\noindent
Special files are particularly interesting because they resolve to OS structures.
One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
% Here are a few examples:
% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---enables the runtime wrapping and rebinding achieved during \sys's transformations.
% TODO: compatibility

A small but powerful set of features is concerned with stream manipulation.
Some of these features are built into the shell;
  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file---which itself can be a FIFO dubbing as input to another command.
% \begin{wrapfigure}{r}{0.15\textwidth}
%   \vspace{-15pt}
%   % \begin{center}
%     % \includegraphics[width=0.48\textwidth]{gull}
%     \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
%   % \end{center}
%   \vspace{-25pt}
% \end{wrapfigure}
More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
% TODO: talk about \ttt{xargs}?

\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  cat * | tee >(wc -l > l.t) | zip > a.z   ($$(p_4)$$)
\end{lstlisting}
\smallskip

@@

\smallskip
\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  read | { calc | square; } >/dev/fd/0     ($$(p_5)$$)
\end{lstlisting}
\smallskip

@@

To conclude, \unix pipelines are surprisingly powerful.
This power is exploited directly by \sys's rewriting passes which re-target the shell rather than an implementation in Java or Go.

@@

\subsection{\sys Design Overview}
\label{bg:overview}

As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
Intuitively,
 (i) the parallel-composition operator exposes task parallelism
% FIXME \kk{I have named that operator level parallelism, which one do we prefer?},
 (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
 (iii) the pipe operator is a candidate for both task and data parallelism, and
% \kk{How does it enable data parallelism?} \nv{added "candidate", helps?}
 (iv)  other composition operators can be viewed as barrier candidates.
To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.

% However, understanding the shell's primitives is not enough
First, \sys needs to understand the distribution characteristics of individual commands used to compose larger programs~\sx{distributability}.
This challenge can be broken down into two parts.
First, understand standard commands part of the shell---that is, built-ins that come with any shell and support its core functions.
One example is the set of commands defined by the \textsc{POSIX} standard~\cite{posix}.
\sys addresses this part by a careful study of a group of commands, identifying a small but clear set of well-understood distributability classes~\sx{cmd}.
This group is already useful-enough to allow the composition of practical pipelines~\sx{micro}.

The second part is understanding the characteristics of commands that fall outside the built-in set, allowing new commands to leverage \sys's power.
New commands are arbitrary programs, but still need to conform to the shell's pipeline interface.
To solve this part, \sys leverages the aforementioned study~\sx{cmd} to define a small but expressive domain-specific language (DSL) for describing a command's key properties with respect to its distributability.
This DSL is aimed towards developers of commands (not users of the shell), who can quickly and easily capture the distributability of the commands they are developing.
% We haven't introduced the notion of flags
% A command developer can only annotate the default behavior;
%   if a new flag is added a few years later, and if that flag alters the command's class, only then a new annotation is needed.

Given a pipeline of commands and their distributability characteristics, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution~\sx{ir}.
It addresses this challenge with a series of rewriting passes that perform dataflow analysis to identify distributable regions.
It converts the AST to a dataflow graph, and iteratively performs graph transformations that expose data parallelism as task parallelism, while preserving the pipeline's correctness.
These transformations are complicated by the shell's pervasive use of dynamic features---\eg environment variables, file input sizes, \etc

To schedule and execute the resulting distributed dataflow graph, \sys needs to solve several challenges~\sx{impl}.
First, it needs to wrap single-node commands with corrective runtime interposition based on their classes.
This is achieved by thin wrappers that wrap all the commands involved in the program, generated using the distributability specification described earlier.
Thin wrappers avoid interference with legacy functionality, as \unix commands contain non-obvious logic and flags added over extended periods of time.

Second, \sys needs to fill-in planing details left blank by the rewriting pass.
This is achieved with \emph{planner activators}, shell expressions that yield (and pass their arguments) back to the planner.
They are inserted by the planner and evaluated purely for their side-effects (\ie yielding to the scheduler) rather than their return value (which is the empty character).
They make clever use of shell features such as command expansion to achieve the latest possible binding, when critical runtime information is available.
% From the point of view of the shell, planner activators are expressed through a higher-order command---similar to \ttt{xargs} and \ttt{time}---that takes other commands as arguments.
Planner activators

% FIXME \kk{I am not exactly sure I understand that :) } \nv{which part?}

The next few sections (\S\ref{distributability}--\ref{impl}) discuss \sys's design and implementation details.
They also outline other, less obvious challenges---such as the synchronization of environment variables, distributed file operation, \etc.
